### 2025-02-05 paper/body_ecml.tex
progress

\begin{equation}\phantomsection\label{eq-obj}{
\text{yloss}(\mathbf{M}_\theta(x),y) + \lambda_{\text{div}} \text{div}({x^\prime},x;\theta) + \lambda_{\text{adv}} \text{advloss}(\mathbf{M}_\theta({x^\prime}),y)
}\end{equation}

@@

A key empirical finding in \cite{altmeyer2024faithful} was that JEMs
tend to do well with respect to the plausibility objective in
Definition~\ref{def-explainability}. If we consider samples drawn from
\(p_\theta(\mathbf{x})\) as counterfactuals, this is an expected
finding, because the JEM objective effectively minimizes the divergence
between the conditional posterior and \(p(\mathbf{x}|\mathbf{y}^{+})\).
To generate samples, \cite{grathwohl2020your} rely on Stochastic
Gradient Langevin Dynamics (SGLD) using an uninformative prior for
initialization. This is where we depart from their methodology: instead
of generating samples through SGLD, we propose using counterfactual
explainers to generate counterfactuals for observed training samples.

This departure allows us to tap into the vast repertoire of explainers
that have been proposed in the literature to meet different desiderata.
Typically, these methods facilitate the imposition of domain and
mutability constraints, for example. In principle, any existing approach
for generating counterfactual explanations is viable, so long as it does
not violate the faithfulness condition.

@@

A reasonable choice for the latter is to define
\(\text{advloss}(\mathbf{M}_\theta({x^\prime}),y;\varepsilon):=\text{yloss}(\mathbf{M}_\theta({x^\prime}),y)\)

>>>>

\begin{equation}\phantomsection\label{eq-obj}{
\text{yloss}(\mathbf{M}_\theta(\mathbf{x}),y) + \lambda_{\text{div}} \text{div}(\mathbf{x},\mathbf{x^\prime};\theta) + \lambda_{\text{adv}} \text{advloss}(\mathbf{M}_\theta(\mathbf{x^\prime}),y)
}\end{equation}

@@

A key empirical finding in \cite{altmeyer2024faithful} was that JEMs
tend to do well with respect to the plausibility objective in
Definition~\ref{def-explainability}. If we consider samples drawn from
\(p_\theta(\mathbf{x})\) as counterfactuals, this is an expected
finding, because the JEM objective effectively minimizes the divergence
between the conditional posterior and \(p(\mathbf{x}|\mathbf{y}^{+})\).
To generate samples, \cite{grathwohl2020your} rely on Stochastic
Gradient Langevin Dynamics (SGLD) using an uninformative prior for
initialization. This is where we depart from their methodology: instead
of generating samples through SGLD, we propose using counterfactual
explainers to generate counterfactuals for observed training samples.
Specifically, we have

\begin{equation}\phantomsection\label{eq-div}{
\text{div}(\mathbf{x},\mathbf{x^\prime};\theta) = \mathcal{E}_\theta(\mathbf{x}) - \mathcal{E}_\theta(\mathbf{x}^\prime)
}\end{equation}

where \(\mathcal{E}_\theta(\cdot)\) denotes the energy function. We
generate samples \(\mathbf{x}^\prime\) by first randomly sampling the
target class \(\mathbf{y}^+ \sim p(\mathbf{y})\) and then generating a
counterfactual explanation for that target, similar to how conditional
sampling can be used for JEMs \cite{grathwohl2020your}. Intuitively, the
gradient of Equation~\ref{eq-div} decreases the energy of observed
training samples (positive samples) while at same time increasing the
energy of counterfactuals (negative samples) \cite{du2019implicit}. As
the generated counterfactuals get more plausible
(Definition~\ref{def-explainability}) over the cause of training, these
two opposing effects gradually balance each out \cite{lippe2024uvadlc}.

The departure from SGLD allows us to tap into the vast repertoire of
explainers that have been proposed in the literature to meet different
desiderata. Typically, these methods facilitate the imposition of domain
and mutability constraints, for example. In principle, any existing
approach for generating counterfactual explanations is viable, so long
as it does not violate the faithfulness condition. Like JEMs
\cite{murphy2022probabilistic}, counterfactual training can be
considered as a form of contrastive representation learning.

@@

Based on our analysis in Section~\ref{sec-lit}, counterfactuals
\(\mathbf{x}^\prime\) can be repurposed as additional training samples
\cite{luu2023counterfactual} or adversarial examples
\cite{freiesleben2022intriguing, pawelczyk2022exploring}. This leaves
some flexibility with respect to the exact choice for
\(\text{advloss}(\cdot)\) in Equation~\ref{eq-obj}. An intuitive
functional form to use, though likely not the only reasonable choice, is
inspired by adversarial training:

\begin{equation}\phantomsection\label{eq-adv}{
\begin{aligned}
\text{advloss}(\mathbf{M}_\theta(\mathbf{x^\prime}),\mathbf{y};\varepsilon)&=\begin{cases}
\text{yloss}(\mathbf{M}_\theta(\mathbf{x^\prime}),\mathbf{y}) & \text{if} \ ||\Delta||_\infty \leq \varepsilon \\
0 & \text{otherwise.}
\end{cases}
\end{aligned}
}\end{equation}

Under this choice we treat the counterfactual \(\mathbf{x}^\prime\) as
an adversarial example iff it is imperceptible, i.e.~the magnitude of
the perturbation of any individual feature is upper-bounded at
\(\varepsilon\).

### 2025-02-05 paper/preprint/paper.tex
progress

A key empirical finding in Altmeyer et al.
(\citeproc{ref-altmeyer2024faithful}{2024}) was that JEMs tend to do
well with respect to the plausibility objective in
Definition~\ref{def-explainability}. If we consider samples drawn from
\(p_\theta(\mathbf{x})\) as counterfactuals, this is an expected
finding, because the JEM objective effectively minimizes the divergence
between the conditional posterior and \(p(\mathbf{x}|\mathbf{y}^{+})\).
To generate samples, Grathwohl et al.
(\citeproc{ref-grathwohl2020your}{2020}) rely on Stochastic Gradient
Langevin Dynamics (SGLD) using an uninformative prior for
initialization. This is where we depart from their methodology: instead
of generating samples through SGLD, we propose using counterfactual
explainers to generate counterfactuals for observed training samples.

This departure allows us to tap into the vast repertoire of explainers
that have been proposed in the literature to meet different desiderata.
Typically, these methods facilitate the imposition of domain and
mutability constraints, for example. In principle, any existing approach
for generating counterfactual explanations is viable, so long as it does
not violate the faithfulness condition.

@@

A reasonable choice for the latter is to define
\(\text{advloss}(\mathbf{M}_\theta({x^\prime}),y;\varepsilon):=\text{yloss}(\mathbf{M}_\theta({x^\prime}),y)\)

@@

@@

@@

>>>>

A key empirical finding in Altmeyer et al.
(\citeproc{ref-altmeyer2024faithful}{2024}) was that JEMs tend to do
well with respect to the plausibility objective in
Definition~\ref{def-explainability}. If we consider samples drawn from
\(p_\theta(\mathbf{x})\) as counterfactuals, this is an expected
finding, because the JEM objective effectively minimizes the divergence
between the conditional posterior and \(p(\mathbf{x}|\mathbf{y}^{+})\).
To generate samples, Grathwohl et al.
(\citeproc{ref-grathwohl2020your}{2020}) rely on Stochastic Gradient
Langevin Dynamics (SGLD) using an uninformative prior for
initialization. This is where we depart from their methodology: instead
of generating samples through SGLD, we propose using counterfactual
explainers to generate counterfactuals for observed training samples.
Specifically, we have

\begin{equation}\phantomsection\label{eq-div}{
\text{div}(\mathbf{x},\mathbf{x^\prime};\theta) = \mathcal{E}_\theta(\mathbf{x}) - \mathcal{E}_\theta(\mathbf{x}^\prime)
}\end{equation}

where \(\mathcal{E}_\theta(\cdot)\) denotes the energy function. We
generate samples \(\mathbf{x}^\prime\) by first randomly sampling the
target class \(\mathbf{y}^+ \sim p(\mathbf{y})\) and then generating a
counterfactual explanation for that target, similar to how conditional
sampling can be used for JEMs
(\citeproc{ref-grathwohl2020your}{Grathwohl et al. 2020}). Intuitively,
the gradient of Equation~\ref{eq-div} decreases the energy of observed
training samples (positive samples) while at same time increasing the
energy of counterfactuals (negative samples)
(\citeproc{ref-du2019implicit}{Du and Mordatch 2020}). As the generated
counterfactuals get more plausible (Definition~\ref{def-explainability})
over the cause of training, these two opposing effects gradually balance
each out (\citeproc{ref-lippe2024uvadlc}{Lippe 2024}).

The departure from SGLD allows us to tap into the vast repertoire of
explainers that have been proposed in the literature to meet different
desiderata. Typically, these methods facilitate the imposition of domain
and mutability constraints, for example. In principle, any existing
approach for generating counterfactual explanations is viable, so long
as it does not violate the faithfulness condition. Like JEMs
(\citeproc{ref-murphy2022probabilistic}{Murphy 2022}), counterfactual
training can be considered as a form of contrastive representation
learning.

@@

Based on our analysis in Section~\ref{sec-lit}, counterfactuals
\(\mathbf{x}^\prime\) can be repurposed as additional training samples
(\citeproc{ref-luu2023counterfactual}{Luu and Inoue 2023}) or
adversarial examples
(\citeproc{ref-freiesleben2022intriguing}{Freiesleben 2022};
\citeproc{ref-pawelczyk2022exploring}{Pawelczyk et al. 2022}). This
leaves some flexibility with respect to the exact choice for
\(\text{advloss}(\cdot)\) in Equation~\ref{eq-obj}. An intuitive
functional form to use, though likely not the only reasonable choice, is
inspired by adversarial training:

\begin{equation}\phantomsection\label{eq-adv}{
\begin{aligned}
\text{advloss}(\mathbf{M}_\theta(\mathbf{x^\prime}),\mathbf{y};\varepsilon)&=\begin{cases}
\text{yloss}(\mathbf{M}_\theta(\mathbf{x^\prime}),\mathbf{y}) & \text{if} \ ||\Delta||_\infty \leq \varepsilon \\
0 & \text{otherwise.}
\end{cases}
\end{aligned}
}\end{equation}

Under this choice we treat the counterfactual \(\mathbf{x}^\prime\) as
an adversarial example iff it is imperceptible, i.e.~the magnitude of
the perturbation of any individual feature is upper-bounded at
\(\varepsilon\).

@@

\bibitem[\citeproctext]{ref-du2019implicit}
Du, Yilun, and Igor Mordatch. 2020. {``Implicit Generation and
Generalization in Energy-Based Models.''}
\url{https://arxiv.org/abs/1903.08689}.

@@

\bibitem[\citeproctext]{ref-lippe2024uvadlc}
Lippe, Phillip. 2024. {``{UvA Deep Learning Tutorials}.''}
\url{https://uvadlc-notebooks.readthedocs.io/en/latest/}.

@@

\bibitem[\citeproctext]{ref-murphy2022probabilistic}
Murphy, Kevin P. 2022. \emph{Probabilistic {Machine Learning}: {An}
Introduction}. {MIT Press}.

### 2025-02-05 paper/sections/method.qmd
progress

$$
\text{yloss}(\mathbf{M}_\theta(x),y) + \lambda_{\text{div}} \text{div}({x^\prime},x;\theta) + \lambda_{\text{adv}} \text{advloss}(\mathbf{M}_\theta({x^\prime}),y)
$$ {#eq-obj}

@@

A key empirical finding in @altmeyer2024faithful was that JEMs tend to do well with respect to the plausibility objective in @def-explainability. If we consider samples drawn from $p_\theta(\mathbf{x})$ as counterfactuals, this is an expected finding, because the JEM objective effectively minimizes the divergence between the conditional posterior and $p(\mathbf{x}|\mathbf{y}^{+})$. To generate samples, @grathwohl2020your rely on Stochastic Gradient Langevin Dynamics (SGLD) using an uninformative prior for initialization. This is where we depart from their methodology: instead of generating samples through SGLD, we propose using counterfactual explainers to generate counterfactuals for observed training samples.

This departure allows us to tap into the vast repertoire of explainers that have been proposed in the literature to meet different desiderata. Typically, these methods facilitate the imposition of domain and mutability constraints, for example. In principle, any existing approach for generating counterfactual explanations is viable, so long as it does not violate the faithfulness condition.

@@

A reasonable choice for the latter is to define $\text{advloss}(\mathbf{M}_\theta({x^\prime}),y;\varepsilon):=\text{yloss}(\mathbf{M}_\theta({x^\prime}),y)$

>>>>

$$
\text{yloss}(\mathbf{M}_\theta(\mathbf{x}),y) + \lambda_{\text{div}} \text{div}(\mathbf{x},\mathbf{x^\prime};\theta) + \lambda_{\text{adv}} \text{advloss}(\mathbf{M}_\theta(\mathbf{x^\prime}),y)
$$ {#eq-obj}

@@

A key empirical finding in @altmeyer2024faithful was that JEMs tend to do well with respect to the plausibility objective in @def-explainability. If we consider samples drawn from $p_\theta(\mathbf{x})$ as counterfactuals, this is an expected finding, because the JEM objective effectively minimizes the divergence between the conditional posterior and $p(\mathbf{x}|\mathbf{y}^{+})$. To generate samples, @grathwohl2020your rely on Stochastic Gradient Langevin Dynamics (SGLD) using an uninformative prior for initialization. This is where we depart from their methodology: instead of generating samples through SGLD, we propose using counterfactual explainers to generate counterfactuals for observed training samples. Specifically, we have

$$
\text{div}(\mathbf{x},\mathbf{x^\prime};\theta) = \mathcal{E}_\theta(\mathbf{x}) - \mathcal{E}_\theta(\mathbf{x}^\prime)
$$ {#eq-div}

where $\mathcal{E}_\theta(\cdot)$ denotes the energy function. We generate samples $\mathbf{x}^\prime$ by first randomly sampling the target class $\mathbf{y}^+ \sim p(\mathbf{y})$ and then generating a counterfactual explanation for that target, similar to how conditional sampling can be used for JEMs [@grathwohl2020your]. Intuitively, the gradient of @eq-div decreases the energy of observed training samples (positive samples) while at same time increasing the energy of counterfactuals (negative samples) [@du2019implicit]. As the generated counterfactuals get more plausible (@def-explainability) over the cause of training, these two opposing effects gradually balance each out [@lippe2024uvadlc].

The departure from SGLD allows us to tap into the vast repertoire of explainers that have been proposed in the literature to meet different desiderata. Typically, these methods facilitate the imposition of domain and mutability constraints, for example. In principle, any existing approach for generating counterfactual explanations is viable, so long as it does not violate the faithfulness condition. Like JEMs [@murphy2022probabilistic], counterfactual training can be considered as a form of contrastive representation learning.

@@

Based on our analysis in @sec-lit, counterfactuals $\mathbf{x}^\prime$ can be repurposed as additional training samples [@luu2023counterfactual] or adversarial examples [@freiesleben2022intriguing;@pawelczyk2022exploring]. This leaves some flexibility with respect to the exact choice for $\text{advloss}(\cdot)$ in @eq-obj. An intuitive functional form to use, though likely not the only reasonable choice, is inspired by adversarial training:

$$
\begin{aligned}
\text{advloss}(\mathbf{M}_\theta(\mathbf{x^\prime}),\mathbf{y};\varepsilon)&=\begin{cases}
\text{yloss}(\mathbf{M}_\theta(\mathbf{x^\prime}),\mathbf{y}) & \text{if} \ ||\Delta||_\infty \leq \varepsilon \\
0 & \text{otherwise.}
\end{cases}
\end{aligned}
$$ {#eq-adv}

Under this choice we treat the counterfactual $\mathbf{x}^\prime$ as an adversarial example iff it is imperceptible, i.e. the magnitude of the perturbation of any individual feature is upper-bounded at $\varepsilon$.
